Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-05-11 3.Thesis/Writeup/2.Manuscript.Rmd
Final version before forking

---
title: "Missing the Point: Non-Convergence in Iterative Imputation Algorithms"
runninghead: Oberman
author:
  name: H. I. Oberman
  num: 1
address:
  num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "Iterative imputation is a popular tool to accommodate missing data. While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence. There is no consensus on how to evaluate the convergence properties of the method. This paper provides insight into identifying non-convergence in iterative imputation algorithms. We conclude that in the case of drawing inference from incomplete data, convergence of the iterative imputation algorithm is often a convenience but not a necessity."
keywords: missing data, iterative imputation, non-convergence, MICE
classoption:
 - Royal
 - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
  bookdown::pdf_book:
    base_format: rticles::sage_article
    keep_tex: yes
---

@@

These remain open questions, since the convergence properties of iterative imputation algorithms have not been systematically studied [@zhu15; @buur18]. There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@taka17]. Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18]. Therefore, algorithmic convergence should be monitored carefully, but this is not straightforward. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13]. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09]. Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms. 
<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\rm inc}$ in dataset $y$. The collection of observed data values in $y$ is denoted by $y_{\rm obs}$; the missing part of $y$ is referred to as $y_{\rm mis}$. For each datapoint in $y_{\rm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm. The state space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values). The collection of $\theta$-values between $t=1$ (summarizing the state space of the algorithm at initialization) and $t=T$ (summarizing the state space for the imputed values) will be referred to as an 'imputation chain'. 
<!-- **[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] ** -->

@@

<!-- **The goal of imputation through e.g. MICE is to obtain a multivariate distribution by iterating over a sequence of univariate imputations. Traditionally, the inspection of the multivariate convergence is done by visually inspecting the univariate chain means and variances over the iterations. The downside to these approaches is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Focusing on the convergence of these outcome parameters may influence this procedure in the sense that the model of evaluation favors the model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. We propose eigenvalue as such a parameter, because it.... ** -->

To illustrate what non-mixing and non-stationarity looks like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2]. Figure \ref{fig:non-conv} displays two scenarios from the example. The graph on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. Each line portrays one imputation chain (i.e., the values of a scalar summary $\theta$ across iterations). The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in $\dot y_{\rm imp, \ell}$.

@@

Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection. After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots. In a traceplot, a scalar summary of the state space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}. The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances. Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18].

Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18, $\S$ 6.5.2]. Moreover, traceplots are typically only used to inspect  univariate $\theta$s. Monitoring univariate summaries of the state space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations. Ideally, we would monitor a multivariate $\theta$.

@@

<!-- **[MERGE with sections above]** Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations. This is typically done through traceplots. In a traceplot, the iteration number is plotted against a certain $\theta$. $\theta$s are scalar summaries of the state space of the algorithm at a specific iteration.  -->

@@

To monitor non-convergence in iterative imputation algorithms, we may  summarize the state space of the algorithm with several $\theta$s. The downside to the current $\theta$s is that they either focus on the univariate state space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent $\theta$ that summarizes the multivariate nature of the data.

We propose $\lambda_{1}$ as such a scalar summary. We define $\lambda_{1}$ as the first eigenvalue of the variance-covariance matrix in the $m$ completed datasets. The eigenvalues of a variance-covariance matrix $S$ summarize the total set of covariances in the data. Let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation $\ell$. $\lambda_{1,\ell}$ then summarizes the largest possible amount of covariance in the $m$ completed datasets {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}. By definition, each $\lambda_{1, \ell}$-value is equal to the variance of the first component that would be obtained by performing principal component analysis on {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}. As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the substantive model of interest, yet still summarizes the multivariate state space of the algorithm. \newline

<!-- **Since convergence of iterative imputation algorithms is in distribution, there is not a unique point at which the algorithm reaches a converged state. We can only evaluate scalar summaries of the multivariate state space of the algorithm, $\theta$s. The recommended scalar summaries to evaluate are chain means and chain variances. Additionally, researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2]. Because of the issues mentioned above, we propose a novel $\theta$ to monitor: the first eigenvalue of the variance-covariance matrix of the completed data. Let $\lambda_{1, \ell} \geq \lambda_2 \geq ... \geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $m$ completed datasets ${y_{\rm obs}, y_{\rm imp}}$. $\lambda_{1, \ell}$ is measure that summarizes the covariances in the completed datasets.**  -->

\noindent In this study, we consider each of the four $\theta$s that were discussed above. Namely, two univariate scalar summaries (chain mean and chain variance), a model-dependent multivariate summary $\hat{Q}$, and the novel model-independent multivariate summary $\lambda_1$.

@@

There are many diagnostic tools to identify non-convergence in iterative (MCMC) algorithms [@broo98; @elad06]. We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity. As recommended by e.g. @cowl96 we will use the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92] to evaluate mixing, and autocorrelation [$AC$; @scha97; @gelm13] to diagnose trending. With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19]. Therefore, we will evaluate the appropriateness of both $\widehat{R}$ versions, in addition to $AC$. Other methods are outside the scope of this study, because they e.g. assume that values within chains represent independent samples, whereas MI only uses the final iteration to impute.

@@

\noindent From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which **approximates** the total variance of $\theta$. $\widehat{R}$ is then obtained as a ratio between the total variance and the within-chain variance:

@@

The $\widehat{R}$-values in panel D are conform our expectations. The adapted version of $\widehat{R}$ depicted here does indicate less signs of non-convergence as the number of iterations goes up. An exception to this general trend is the 'dip' around $t=3$. As diagnostic tool, this may lead to the incorrect conclusion that the non-convergence scenario has superior convergence compared to the typical scenario. We can explain this as a downside of the low number of iteration in iterative imputation: the adapted version of $\widehat{R}$ can only be completely employed if the number of iterations is at least four. Otherwise, it is not possible to perform all three transformations to the chains of $\theta$-values. For low values of $t$, the $\widehat{R}$-values conform Vehtari et al. are, therefore, more similar to the original $\widehat{R}$ by Gelman and Rubin. \newline
<!-- However, the superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.  -->

\noindent In short, both non-convergence diagnostics have a version that may be appropriate for iterative imputation algorithms. We will compute $\widehat{R}$ conform @veht19, and calculate $AC$ manually. The diagnostics will be applied to the four types of scalar summaries $\theta$ described above: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix. We evaluate the performance of the eight sets of diagnostic methods (two diagnostics; four $\theta$s) through simulation.

@@

The aim of the simulation study is to determine to what extend non-convergence affects iterative imputation algorithms. Specifically, we are interested in the validity of inferences drawn from incomplete data using the MICE algorithm. And, whether invalid inferences due to non-convergence may be detected using eight sets of diagnostic methods. We will conclude that the algorithm is sufficiently converged for practical purposes when the estimated quantities of scientific interest $\bar{Q}$ are unbiased and confidence-valid estimates of ${Q}$. In contrast to this, we defined *approximate* convergence as the most converged state that the algorithm can obtain. Continued iteration after reaching approximate convergence does not yield better mixing or stationarity (indicated by non-improving $\widehat{R}$- and $AC$-values). Continued iteration after obtaining valid inferences may lead to a more converged state of the algorithm, but not better estimates.

@@

To induce non-convergence, we consider two sets of simulation conditions: 'incompleteness' and 'early stopping'. Incompleteness refers to the  proportion of incomplete cases in dataset $y$, $p_{\rm inc}$. Early stopping is defined by the number of iterations in the imputation algorithm, $T$.

@@

The incompleteness conditions are chosen to reflect the difficulty of the missingness problem. The underlying assumption is that a low proportion of incomplete cases $p_{\rm inc}$ leads to quick algorithmic convergence, since there is a lot of information in the observed data. Higher $p_{\rm inc}$ yields slower convergence---unless there is so little information in $y_{\rm obs}$ that this is outweighed by the random component in the imputation algorithm. Then, convergence to a stable but highly variable state may be reached instantly. Still, we expect that incompleteness conditions with higher $p_{\rm inc}$ will result in more signs of non-convergence.

@@

<!-- **[REMOVE?] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $m$ completed datasets, and finally pooling the $m$ estimates $\hat{Q}_\ell$ to obtain a single $\bar{Q}$ for each $Q$. We try to predict the effect of non-convergence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by applying these methods to several scalar summaries of the state space of the algorithm, $\theta$s.** -->

@@

3. We hypothesize that the recommended thresholds to diagnose non-convergence with $\widehat{R}$ ($\widehat{R} > 1.2$, $\widehat{R} > 1.1$, and $\widehat{R} > 1.01$) may be too stringent for iterative imputation applications. In an empirical study, where $\widehat{R}$ was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold $\widehat{R}>1.2$. Yet, scientific estimates were insensitive to continued iteration from $t=5$ onward [@lace07]. We, therefore, suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates. In contrast to this, it may occur that signs of non-convergence are under-estimated by $\widehat{R}$, in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed [@broo98, p. 437]. In  `mice`, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values. In practice, we do not expect this to cause problems for identifying non-convergence with $\widehat{R}$.

@@

In each simulation repetition, we *ampute* the complete dataset conform the incompleteness conditions. The proportion of incomplete cases $p_{\rm inc}$ varies between 5% and 95%, in incremental steps ($p_{\rm inc}$ = .05, .25, .50, .75, .95). We consider all possible univariate and multivariate patterns of missingness, conform a 'missing completely at random' missingness mechanism [MCAR; @rubin87]. I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on $p_{\rm inc}$. We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\rm obs}, y_{\rm mis}$} for each of the five incompleteness conditions.

We impute the missing values in {$y_{\rm obs}, y_{\rm mis}$} according to the early stopping conditions. That is, we vary the number of iterations in the imputation algorithm between one and one hundred ($T = 1,2,.,100$). All imputation procedures are performed using `mice` [function `mice()`; @mice], with Bayesian linear regression imputation, and five imputation chains ($m=5$). This results in five sets of imputations for each of the one hundred early stopping conditions, $y_{\rm imp, \ell}$.

Subsequently, we obtain the completed data per imputation, {$y_{\rm obs}, y_{\rm imp, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice]. We estimate univariate $Q$s directly from the completed data, by averaging across imputations to obtain $\bar{Q}$. Estimates of multivariate $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R]. We pool the resulting estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform @vink14.

@@

We consider eight sets of diagnostic methods to identify non-convergence, by applying $\widehat{R}$ and $AC$ on four different $\theta$s. Univariate $\theta$s are obtained from $y_{\rm imp, \ell}$. The model-independent multivariate $\theta$, $\lambda_{1}$, is obtained from {$y_{\rm obs}, y_{\rm imp, \ell}$}. The model-dependent multivariate $\theta$ is one of the quantities of scientific interest: the estimated regression coefficients in {$y_{\rm obs}, y_{\rm imp, \ell}$}. We apply the two non-convergence diagnostics on the $\theta$s by implementing the adapted version of $\widehat{R}$ and by manually programming the $AC$ function.

@@

Our results show the performance of the MICE algorithm under different conditions of incompleteness and early stopping. For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias. For univariate scientific estimands ($Q=\mu$ and $Q=\sigma$) we observe the largest bias in the outcome variable $Y$. And for the regression coefficients ($Q=\beta$) the bias is most pronounced in $\beta_1$ (the effect of $X_1$ on $Y$). Since there is just one estimate for $Q=r^2$ to evaluate, we consider $Q=\mu_Y, \sigma_Y,r^2, \beta_1$. In the figures, we present all incompleteness conditions, but only early stopping conditions where $1 \leq T\leq50$. That is, the results are more or less stable for conditions where $T \geq 30$. Full results are available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/).

@@

Figure \ref{fig:Qs} displays the influence of incompleteness and early stopping on the estimated quantities of scientific interest, $Q$s. In the first row of the figure, we see the bias in the estimated descriptive statistics ($Q=\mu_Y$ and $Q=\sigma_Y$). The second row shows the bias in the multivariate scientific estimands ($Q=r^2$ and $Q=\beta_1$). The third row consists of the coverage rate (CR) and confidence interval width (CIW) of the estimated regression coefficient ($Q=\beta_1$).

@@

$AC$-values are systematically higher for this $\theta$ than for other scalar summaries. The $AC$-values also taper off at a later point in the iterations. This suggests improving stationarity up-to ten, or even thirty iterations, depending on the missingness proportion. Using the threshold, we diagnose non-convergence at some point in all incompleteness conditions. However, this occurs only after reasonably decreasing $AC$-values are obtained---again suggesting that non-convergence is incorrectly diagnosed. \newline

>>>>

---
title: "Missing the Point: Non-Convergence in Iterative Imputation Algorithms"
runninghead: Oberman
author:
  name: H. I. Oberman
  num: 1
address:
  num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "Iterative imputation is a popular tool to accommodate missing data. While it is widely accepted that valid inferences can be obtained with this technique, these inferences all rely on algorithmic convergence. There is no consensus on how to evaluate the convergence properties of the method. This paper provides insight into identifying non-convergence in iterative imputation algorithms. Our study found that---in the cases considered---inferential validity was achieved after five to ten iterations, much earlier than indicated by the $\\widehat{R}$ and AC diagnostics. Of course, it never hurts to iterate longer, but such calculations hardly bring added value."
keywords: missing data, iterative imputation, non-convergence, MICE
classoption:
 - Royal
 - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
  bookdown::pdf_book:
    base_format: rticles::sage_article
    keep_tex: yes
---

@@

These remain open questions since the convergence properties of iterative imputation algorithms have not been systematically studied [@buur18]. There is no scientific consensus on how to evaluate the convergence of imputation algorithms [@zhu15; @taka17]. Moreover, the behavior of such algorithms under certain default imputation models (e.g., 'predictive mean matching') is an entirely open question [@murr18]. Therefore, algorithmic convergence should be monitored carefully, but this is not straightforward. Iterative imputation algorithms such as MICE are special cases of Markov chain Monte Carlo (MCMC) methods. In MCMC methods, convergence is not from a scalar to a point, but from one distribution to another. The values generated by the algorithm (e.g., imputed values) will vary even after convergence [@gelm13]. Since MCMC algorithms do not reach a unique point at which convergence is established, diagnostic methods may only identify signs of *non*-convergence [@hoff09]. Several non-convergence diagnostics exist, but it is not known whether these are appropriate for iterative imputation algorithms. 
<!-- "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

@@

Let $y$ denote an $n \times k$ matrix containing the data values on $k$ variables for all $n$ units in a sample. The data value of unit $i$ ($i = 1, 2, \dots, n$) on variable $j$ ($j = 1, 2, \dots, k$) may be either observed or missing. The number of units $i$ with at least one missing value, divided by the total number of units $n$, is called the proportion of incomplete cases $p_{\rm inc}$ in dataset $y$. The collection of observed data values in $y$ is denoted by $y_{\rm obs}$; the missing part of $y$ is referred to as $y_{\rm mis}$. For each datapoint in $y_{\rm mis}$, we sample $m \times T$ plausible values, where $m$ is the number of imputations ($\ell = 1, 2, \dots, m$) and $T$ is the number of iterations ($t = 1, 2, \dots, T$) in the imputation algorithm. The state-space of the algorithm at a certain iteration $t$ may be summarized by a scalar summary $\theta$ (e.g., the average of the imputed values). The collection of $\theta$-values between $t=1$ (summarizing the state-space of the algorithm at initialization) and $t=T$ (summarizing the state-space for the imputed values) will be referred to as an 'imputation chain'. 
<!-- **[If not defined elsewhere, add: $\theta$s are scalar summaries of the algorithm at a certain iteration (e.g., chain means; the average of the imputed values in each imputation chain).] ** -->

@@

<!-- **The goal of imputation through e.g. MICE is to obtain a multivariate distribution by iterating over a sequence of univariate imputations. Traditionally, the inspection of the multivariate convergence is done by visually inspecting the univariate chain means and variances over the iterations. The downside to these approaches is that they either focus on the univariate state-space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Focusing on the convergence of these outcome parameters may influence this procedure in the sense that the model of evaluation favors the model of interest. Ideally, one would like to evaluate a model-independent parameter that summarizes the multivariate nature of the data. We propose eigenvalue as such a parameter, because it.... ** -->

To illustrate what non-mixing and non-stationarity look like in iterative imputation algorithms, we reproduce an example from van Buuren [-@buur18, $\S$ 6.5.2]. Figure \ref{fig:non-conv} displays two scenarios from the example. The graph on the left-hand side of the figure shows typical convergence of an iterative imputation algorithm. The right-hand side displays pathological non-convergence, induced by purposefully mis-specifying the imputation model. Each line portrays one imputation chain (i.e., the values of a scalar summary $\theta$ across iterations). The $\theta$ depicted here is the 'chain mean' of variable $j$, which is defined as the average of variable $j$ in $\dot y_{\rm imp, \ell}$.

@@

Currently, the recommended practice for evaluating the convergence of the MICE algorithm is through visual inspection. After running the imputation algorithm for a certain number of iterations, researchers are encouraged to produce traceplots. In a traceplot, a scalar summary of the state-space of the algorithm $\theta$ is plotted against the iteration number, as depicted in Figure \ref{fig:non-conv}. The default scalar summaries to inspect for the MICE algorithm are chain means and chain variances. Non-convergence is diagnosed if the imputation chains are not freely intermingled with one another, or if the chains show definite trends [@buur18].

Identifying non-convergence by inspecting traceplots may be undesirable for several reasons: 1) it may be challenging to the untrained eye, 2) only severely pathological cases of non-convergence may be diagnosed, and 3) there is not an objective measure that quantifies convergence [@buur18, $\S$ 6.5.2]. Moreover, traceplots are typically only used to inspect  univariate $\theta$s. Monitoring univariate summaries of the state-space of the algorithm may be insufficient because MICE is not only concerned with a single column, but the entire multivariate distribution of the imputations. Ideally, we would monitor a multivariate $\theta$.

@@

<!-- **[MERGE with sections above]** Convergence has historically been inspected visually, by monitoring the imputation algorithm over iterations. This is typically done through traceplots. In a traceplot, the iteration number is plotted against a certain $\theta$. $\theta$s are scalar summaries of the state-space of the algorithm at a specific iteration.  -->

@@

To monitor non-convergence in iterative imputation algorithms, we may  summarize the state-space of the algorithm with several $\theta$s. The downside to the current $\theta$s is that they either focus on the univariate state-space, or primarily track the change over the iterations of a multivariate outcome conform the scientific model of interest. Ideally, one would like to evaluate a model-independent $\theta$ that summarizes the multivariate nature of the data.

We propose $\lambda_{1}$ as such a scalar summary. We define $\lambda_{1}$ as the first eigenvalue of the variance-covariance matrix in the $m$ completed datasets. The eigenvalues of a variance-covariance matrix $S$ summarize the total set of covariances in the data. Let $\lambda_{1, \ell} \geq \lambda_{2, \ell} \geq ... \geq \lambda_{j, \ell}$ be the eigenvalues of $S_\ell$ in each imputation $\ell$. $\lambda_{1,\ell}$ then summarizes the largest possible amount of covariance in the $m$ completed datasets {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}. By definition, each $\lambda_{1, \ell}$-value is equal to the variance of the first component that would be obtained by performing principal component analysis (PCA) on {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}. As scalar summary, $\lambda_{1}$ has the appealing property that it is not dependent on the substantive model of interest, yet still summarizes the multivariate state-space of the algorithm.

<!-- **Since convergence of iterative imputation algorithms is in distribution, there is not a unique point at which the algorithm reaches a converged state. We can only evaluate scalar summaries of the multivariate state-space of the algorithm, $\theta$s. The recommended scalar summaries to evaluate are chain means and chain variances. Additionally, researchers may "monitor some statistic of scientific interest" [@buur18, $\S$ 6.5.2]. Because of the issues mentioned above, we propose a novel $\theta$ to monitor: the first eigenvalue of the variance-covariance matrix of the completed data. Let $\lambda_{1, \ell} \geq \lambda_2 \geq ... \geq \lambda_j$ be the eigenvalues of $\Sigma$ in each of the $m$ completed datasets ${y_{\rm obs}, y_{\rm imp}}$. $\lambda_{1, \ell}$ is measure that summarizes the covariances in the completed datasets.**  -->

In this study, we consider each of the four $\theta$s that were discussed above. Namely, two univariate scalar summaries (chain mean and chain variance), a model-dependent multivariate summary $\hat{Q}$, and the novel model-independent multivariate summary $\lambda_1$.

@@

There are many diagnostic tools to identify non-convergence in iterative (MCMC) algorithms [@broo98; @elad06]. We consider only two of them that may be appropriate for imputation algorithms---one to monitor signs of non-mixing, and one for non-stationarity. As recommended by e.g. @cowl96 we will use the potential scale reduction factor $\widehat{R}$ ['Gelman-Rubin statistic'; @gelm92] to evaluate mixing, and autocorrelation [$AC$; @scha97; @gelm13] to diagnose trending. With a recently proposed adaptation, $\widehat{R}$ might also serve to diagnose non-stationarity, but this has not yet been thoroughly investigated [@veht19]. Therefore, we will evaluate the appropriateness of both $\widehat{R}$ versions, in addition to $AC$. Other methods are outside the scope of this study because they e.g. assume that values within chains represent independent samples, whereas MI only uses the final iteration to impute.

@@

\noindent From the between- and within-chain variances we compute a weighted average, $\widehat{\operatorname{var}}^{+}$, which over-estimates the total variance of $\theta$. $\widehat{R}$ is then obtained as a ratio between this total variance and the within-chain variance:

@@

The $\widehat{R}$-values in panel D are conform our expectations. The adapted version of $\widehat{R}$ depicted here does indicate less signs of non-convergence as the number of iterations goes up. An exception to this general trend is the 'dip' around $t=3$. As diagnostic tool, this may lead to the incorrect conclusion that the non-convergence scenario has superior convergence compared to the typical scenario. We can explain this as a downside of the low number of iteration in iterative imputation: the adapted version of $\widehat{R}$ can only be completely employed if the number of iterations is at least four. Otherwise, it is not possible to perform all three transformations to the chains of $\theta$-values. For low values of $t$, the $\widehat{R}$-values conform Vehtari et al. are, therefore, more similar to the original $\widehat{R}$ by Gelman and Rubin.
<!-- However, the superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$.  -->

In short, both non-convergence diagnostics have a version that may be appropriate for iterative imputation algorithms. We will compute $\widehat{R}$ conform @veht19, and calculate $AC$ manually. The diagnostics will be applied to the four types of scalar summaries $\theta$ described above: chain means, chain variances, a quantity of scientific interest, and the first eigenvalue of the variance-covariance matrix. We evaluate the performance of the eight sets of diagnostic methods (two diagnostics; four $\theta$s) through simulation.

@@

The simulation study aims to determine to what extend non-convergence affects iterative imputation algorithms. Specifically, we are interested in the validity of inferences drawn from incomplete data using the MICE algorithm. And, whether invalid inferences due to non-convergence may be detected using eight sets of diagnostic methods. We will conclude that the algorithm is sufficiently converged for practical purposes when the estimated quantities of scientific interest $\bar{Q}$ are unbiased and confidence-valid estimates of ${Q}$. In contrast to this, we defined *approximate* convergence as the most converged state that the algorithm can obtain. Continued iteration after reaching approximate convergence does not yield better mixing or stationarity (indicated by non-improving $\widehat{R}$- and $AC$-values). Continued iteration after obtaining valid inferences may lead to a more converged state of the algorithm, but not better estimates.

@@

To induce non-convergence, we consider two sets of simulation conditions: 'missingness' and 'early stopping'. Missingness refers to the severity of the missing data problem, which we determine by the proportion of incomplete cases in dataset $y$, $p_{\rm inc}$. Early stopping is defined by the number of iterations in the imputation algorithm, $T$.

@@

The missingness conditions are chosen to reflect the difficulty of the missingness problem. The underlying assumption is that a low proportion of incomplete cases $p_{\rm inc}$ leads to quick algorithmic convergence, since there is a lot of information in the observed data. Higher $p_{\rm inc}$ yields slower convergence---unless there is so little information in $y_{\rm obs}$ that this is outweighed by the random component in the imputation algorithm. Then, convergence to a stable but highly variable state may be reached instantly. Still, we expect that missingness conditions with higher $p_{\rm inc}$ will result in more signs of non-convergence.

@@

<!-- **[REMOVE?] $Q$s are the targets that will be estimated in each simulation repetition, after first *amputing* the complete data with varying missingness proportions, then imputing the missingness with a varying number of iterations, then estimating the $Q$s on the $m$ completed datasets, and finally pooling the $m$ estimates $\hat{Q}_\ell$ to obtain a single $\bar{Q}$ for each $Q$. We try to predict the effect of non-convergence (i.e., the difference between $Q$ and $\bar{Q}$) with the convergence diagnostics $\widehat{R}$ and $AC$, by applying these methods to several scalar summaries of the state-space of the algorithm, $\theta$s.** -->

@@

3. We hypothesize that the recommended thresholds to diagnose non-convergence with $\widehat{R}$ ($\widehat{R} > 1.2$, $\widehat{R} > 1.1$, and $\widehat{R} > 1.01$) may be too stringent for iterative imputation applications. In an empirical study, where $\widehat{R}$ was used to inform the required imputation chain length, it took as many as 50 iterations to overcome the conventional non-convergence threshold $\widehat{R}>1.2$. Yet, scientific estimates were insensitive to continued iteration from $t=5$ onward [@lace07]. We, therefore, suspect that $\widehat{R}$ may over-estimate signs of non-convergence in iterative imputation algorithms, compared to the validity of estimates. In contrast to this, signs of non-convergence can be under-estimated by $\widehat{R}$, in exceptional cases where the initial values of the algorithm are not appropriately over-dispersed [@broo98, p. 437]. In  `mice`, initial values are chosen randomly from the observed data, hence we cannot be certain of over-dispersion in the initial values. In practice, we do not expect this to cause problems for identifying non-convergence with $\widehat{R}$.

@@

In each simulation repetition, we *ampute* the complete dataset conform the missingness conditions. The proportion of incomplete cases $p_{\rm inc}$ varies between 5% and 95%, in incremental steps ($p_{\rm inc}$ = .05, .25, .50, .75, .95). We consider all possible univariate and multivariate patterns of missingness, conform a 'missing completely at random' missingness mechanism [MCAR; @rubin87]. I.e., the probability to be missing is the same for all $N \times k$ cells in $y$, conditional on $p_{\rm inc}$. We use the 'mice' package [function `mice::ampute()`; @mice] to obtain an incomplete dataset {$y_{\rm obs}, y_{\rm mis}$} for each of the five missingness conditions.

We impute the missing values in {$y_{\rm obs}, \dot y_{\rm mis}$} according to the early stopping conditions. That is, we vary the number of iterations in the imputation algorithm between one and one hundred ($T = 1,2,.,100$). All imputation procedures are performed using `mice` [function `mice()`; @mice], with Bayesian linear regression imputation, and five imputation chains ($m=5$). This results in $m=5$ sets of imputations for each of the one hundred early stopping conditions, $\dot y_{\rm imp, \ell}$.

Subsequently, we obtain the completed data per imputation, {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}, by combining the imputed data with the observed data [function `mice::complete()`; @mice]. We estimate univariate $Q$s directly from the completed data, by averaging across imputations to obtain $\bar{Q}$. Estimates of multivariate $Q$s are obtained after performing multiple linear regression [function `stats::lm()`, @R]. We pool the resulting estimates, $\hat{Q}_\ell$, into a single $\bar{Q}$ conform @vink14.

@@

We consider eight sets of diagnostic methods to identify non-convergence, by applying $\widehat{R}$ and $AC$ on four different $\theta$s. Univariate $\theta$s are obtained from $\dot y_{\rm imp, \ell}$. The model-independent multivariate $\theta$, $\lambda_{1}$, is obtained from {$y_{\rm obs}, \dot y_{\rm imp, \ell}$} as the variance in the first PCA component [function `stats::princomp`; R]. The model-dependent multivariate $\theta$ is one of the quantities of scientific interest: the estimated regression coefficients in {$y_{\rm obs}, \dot y_{\rm imp, \ell}$}. We apply the two non-convergence diagnostics on the $\theta$s by implementing the adapted version of $\widehat{R}$ and by manually programming the $AC$ function.

@@

Our results show the performance of the MICE algorithm under different conditions of missingness and early stopping. For reasons of brevity, we only discuss results for the worst-performing estimates in terms of bias. For univariate scientific estimands ($Q=\mu$ and $Q=\sigma$) we observe the largest bias in the outcome variable $Y$. And for the regression coefficients ($Q=\beta$) the bias is most pronounced in $\beta_1$ (the effect of $X_1$ on $Y$). Since there is just one estimate for $Q=r^2$ to evaluate, we consider $Q=\mu_Y, \sigma_Y,r^2, \beta_1$. In the figures, we present all missingness conditions, but only early stopping conditions where $1 \leq T\leq50$. That is, the results are more or less stable for conditions where $T \geq 30$. Full results are available from [github.com/gerkovink/shinyMice](https://github.com/gerkovink/shinyMice/tree/master/3.Thesis/).

@@

Figure \ref{fig:Qs} displays the influence of missingness and early stopping on the estimated quantities of scientific interest, $Q$s. In the first row of the figure, we see the bias in the estimated descriptive statistics ($Q=\mu_Y$ and $Q=\sigma_Y$; panels A and B). The second row shows the bias in the multivariate scientific estimands ($Q=r^2$ and $Q=\beta_1$; C and D). The third row consists of the coverage rate (CR) and confidence interval width (CIW) of the estimated regression coefficient ($Q=\beta_1$; E and F).

@@

$AC$-values are systematically higher for this $\theta$ than for other scalar summaries. The $AC$-values also taper off at a later point in the iterations. This suggests improving stationarity up-to ten, or even thirty iterations, depending on the missingness proportion. Using the threshold, we diagnose non-convergence at some point in all missingness conditions. However, this occurs only after reasonably decreasing $AC$-values are obtained---again suggesting that non-convergence is incorrectly diagnosed.
